---
bibliography: cluster_random.bib
output:
  html_document:
    toc: true
    theme: journal
    includes:
        after_body: ../linking_script.html
---

<style>
div.hidecode + pre {display: none}
</style>

<script>
doclick=function(e){
e.nextSibling.nextSibling.style.display="block";
}
</script>


<!-- title: "10 Things You Need To Know About Cluster Randomization" -->

```{r,echo=FALSE, message=FALSE}
rm(list = ls())
library(knitr)
library(lme4)
```

# 1 What clustering is

Cluster randomized[^1] experiments allocate treatments to groups, but measure
outcomes at the level of the individuals that compose the groups. It is this
divergence between the level at which the intervention is assigned and the level
at which outcomes are defined that classifies an experiment as cluster
randomized.

[^1]: This guide was originally written by Jake Bowers and Ashlea Rundlett (22
Nov 2014). Updates made by Jasper Cooper (25 Aug 2015) and Anna Wilke (Mar
2022).

Consider a study that randomly assigns villages to receive different development
programs, where the well-being of households in the village is the outcome of
interest. This is a clustered design because, while the treatment is assigned to
the village as a whole, we are interested in outcomes defined at the household
level. Or consider a study that randomly assigns certain households to receive
different get-out-the-vote messages, where we are interested in the voting
behavior of individuals. Because the unit of assignment for the message is the
household, but the outcome is defined as individual behavior, this study is
cluster randomized.

Now consider a study in which villages are selected at random, and 10 people
from each village are assigned to treatment or control, and we measure the
well-being of those individuals. In this case, the study is not cluster
randomized, because the level at which treatment is assigned and the level at
which outcomes are defined is the same. Suppose that a study randomly assigned
villages to different development programs and then measured social cohesion in
the village. Though it contains the same randomization procedure as our first
example, it is not clustered because we are interested here in village-level
outcomes: the level of treatment assignment and of outcome measurement is the
same.

Clustering matters for two main reasons. On the one hand, clustering reduces the
amount of information in an experiment because it restricts the number of ways
that the treatment and control groups can be composed, relative to randomization
at the individual level. If this fact is not taken into account, we often
underestimate the variance in our estimator, leading to over-confidence in our
the results of the study. On the other hand, clustering raises the question of
how to combine information from different parts of the same experiment into one
quantity. Especially when clusters are not of equal sizes and the potential
outcomes of units within them are very different, conventional estimators will
systematically produce the wrong answer due to bias. However, several approaches
at the design and analysis phases can overcome the challenges posed by cluster
randomized designs.


# 2 Why clustering can matter I: information reduction

We commonly think of the information contained in studies in terms of the sample
size and the characteristics of the units within the sample. Yet two studies
with exactly the same sample size and the same participants could in theory
contain very different amounts of information depending on whether units are
clustered. This will greatly affect the precision of the inferences we make
based on the studies.

Imagine an impact evaluation with 10 people, where 5 are assigned to the
treatment group and 5 to control. In one version of the experiment, the
treatment is assigned to individuals - it is not cluster randomized. In another
version of the experiment, the 5 individuals with black hair and the 5
individuals with some other color of hair are assigned to treatment as a group.
This design has two clusters: 'black hair' and 'other color'.

A simple application of the *n* choose *k* rule shows why this matters. In the
first version, the randomization procedure allows for 252 different combinations
of people as the treatment and control groups. However, in the second version,
the randomization procedure assigns all the black-haired subjects either to
treatment or to control, and thus only ever produces two ways of combining
people to estimate an effect.

Throughout this guide, we will illustrate points using examples written in `R`
code. You can copy and paste this into your own `R` program to see how it works.

<div class="hidecode" onclick="doclick(this);">[Click to show code]</div>
```{r,cache=T}
set.seed(12345)
# Define the sample average treatment effect (SATE)
treatment_effect     <- 1
# Define the individual ids (i)
person               <- 1:10
# Define the cluster indicator (j)
hair_color           <- c(rep("black",5),rep("brown",5))
# Define the control outcome (Y0)
outcome_if_untreated <- rnorm(n = 10)
# Define the treatment outcome (Y1)
outcome_if_treated   <- outcome_if_untreated + treatment_effect

# Version 1 - Not cluster randomized
# Generate all possible non-clustered assignments of treatment (Z)
non_clustered_assignments <- combn(x = unique(person),m = 5)
# Estimate the treatment effect
treatment_effects_V1 <-
     apply(
          X = non_clustered_assignments,
          MARGIN = 2,
          FUN = function(assignment) {
               treated_outcomes   <- outcome_if_treated[person %in% assignment]
               untreated_outcomes <- outcome_if_untreated[!person %in% assignment]
               mean(treated_outcomes) - mean(untreated_outcomes)
          }
     )
# Estimate the true standard error
standard_error_V1 <- sd(treatment_effects_V1)
# Plot the histogram of all possible estimates of the treatment effect
hist(treatment_effects_V1,xlim = c(-1,2.5),breaks = 20)
```

<div class="hidecode" onclick="doclick(this);">[Click to show code]</div>
```{r,cache=T}
# Version 2 - Cluster randomized
# Generate all possible assignments of treatment when clustering by hair color (Z)
clustered_assignments     <- combn(x = unique(hair_color),m = 1)
# Estimate the treatment effect
treatment_effects_V2 <-
     sapply(
          X = clustered_assignments,
          FUN = function(assignment) {
               treated_outcomes   <- outcome_if_treated[person %in% person[hair_color==assignment]]
               untreated_outcomes <- outcome_if_untreated[person %in% person[!hair_color==assignment]]
               mean(treated_outcomes) - mean(untreated_outcomes)
          }
     )
# Estimate the true standard error
standard_error_V2 <- sd(treatment_effects_V2)
# Plot the histogram of all possible estimates of the treatment effect
hist(treatment_effects_V2,xlim = c(-1,2.5),breaks = 20)
```

As the histograms show, clustering provides a much 'coarser' view of the
treatment effect. Irrespective of the number of times one randomizes
the treatment and of the number of subjects one has, in a clustered
randomization procedure the number of possible estimates of the treatment effect
will be strictly determined by the number of clusters assigned to the different
treatment conditions. This has important implications for the standard error of
the estimator.

<div class="hidecode" onclick="doclick(this);">[Click to show code]</div>
```{r}
# Compare the standard errors
kable(round(data.frame(standard_error_V1,standard_error_V2),2))
```

While the sampling distribution for the non-clustered estimate of the treatment
effect has a standard error of about .52, that of the clustered estimate is more
than twice as large, at 1.13. Recall that the data going into both studies is
identical, the only difference between the studies resides in the way that the
treatment assignment mechanism reveals the information.

Related to this issue of information is the question of how much units in our
study vary within and between clusters. Two cluster randomized studies with
$J=10$ villages and $n_j=100$ people per village may have different information
about the treatment effect on individuals if, in one study, differences
*between* villages are much greater than the differences in outcomes *within*
them. If, say, all of the individuals in any village acted exactly the same but
different villages showed different outcomes, then we would have on the order of
10 pieces of information: all of the information about causal effects in that
study would be at the village level. Alternatively, if the individuals within a
village acted more or less independently of each other, then we would have on
the order of 10 $\times$ 100=1000 pieces of information.

We can formalize the idea that the highly dependent clusters provide less
information than the highly independent clusters with the **intracluster
correlation coefficient**. For a given variable, $y$, units $i$ and clusters
$j$, we can write the intracluster correlation coefficient as follows:

$$ \text{ICC} \equiv \frac{\text{variance between clusters in } y}{\text{total variance in } y} \equiv \frac{\sigma_j^2}{\sigma_j^2 + \sigma_i^2} $$

where $\sigma_i^2$ is the variance between units in the population and $\sigma_j^2$ is the variance between outcomes defined at the cluster level. @kish1965survey uses this description of dependence to define his idea of the “effective N” of a study (in the sample survey context, where samples may be clustered):

$$\text{effective N}=\frac{N}{1+(n_j -1)\text{ICC}}=\frac{Jn}{1+(n-1)\text{ICC}},$$

where the second term follows if all of the clusters are the same size ($n_1 \ldots n_J \equiv n$).

If 200 observations arose from 10 clusters with 20 individuals within each
cluster, where ICC = .5, such that 50% of the variation could be attributed to
cluster-to-cluster differences (and not to differences within a cluster), Kish’s
formula would suggest that we have an effective sample size of about 19
observations, instead of 200.

As the foregoing discussion suggests, clustering reduces information most when
it a) greatly restricts the number of ways in which subjects can be assigned to
treatment and control groups, and b) produces units whose outcomes are strongly
related to the cluster they are a member of (i.e. when it increases the ICC).

# 3 What to do about information reduction

One way to limit the extent to which clustering reduces the information
contained in our design would be to form clusters in such a way that the
intracluster correlation is low. For example, if we were able to form clusters
randomly, there would be no systematic relationship between units within a
cluster. Hence, using cluster random assignment to assign these randomly formed
clusters to experimental conditions would not result in any information
reduction. Yet, opportunities to create clusters are typically rare. In most
cases, we  have to rely on naturally formed clusters (e.g., villages,
classrooms, cities). In most scenarios that involve clustered assignment, we
hence will have to make sure that our estimates of uncertainty about treatment
effects correctly reflect the resulting information loss from clustering.

In order to characterize our uncertainty about treatment effects, we typically
want to calculate a standard error: an estimate of how much our treatment effect
estimates *would* vary, were we able to repeat the experiment a very large
number of times and, for each repetition, observe units in their resulting
treated or untreated state.

However, we are never able to observe the true standard error of an estimator,
and therefore must use statistical procedures to infer this unknown quantity.
Conventional methods for calculating standard errors do not take into account
clustering. Thus, in order to avoid over-confidence in experimental findings, we
need to modify the way in which we calculate uncertainty estimates.

In this section we limit our attention to so-called 'design-based' approaches to
calculating the standard error. In the design-based approach we simulate
repetitions of the experiment to derive and check ways of characterizing the
variance of the estimate of the treatment effect, accounting for clustered
randomization. We contrast these with 'model-based' approaches further on in the
guide. In the model-based approach we state that the outcomes were generated
according to a probability model and that the cluster-level relationships also
follow a probability model.

To begin, we will create a function which simulates a cluster randomized
experiment with fixed intracluster correlation, and use it to simulate some data
from a simple cluster-randomized design.

<div class="hidecode" onclick="doclick(this);">[Click to show code]</div>
```{r,cache=T}
make_clustered_data <- function(J = 10, n = 100, treatment_effect = .25, ICC = .1){
     ## Inspired by Mathieu et al, 2012, Journal of Applied Psychology
     if (J %% 2 != 0 | n %% 2 !=0) {
          stop(paste("Number of clusters (J) and size of clusters (n) must be even."))
     }
     Y0_j         <- rnorm(J,0,sd = (1 + treatment_effect) ^ 2 * sqrt(ICC))
     fake_data    <- expand.grid(i = 1:n,j = 1:J)
     fake_data$Y0 <- rnorm(n * J,0,sd = (1 + treatment_effect) ^ 2 * sqrt(1 - ICC)) + Y0_j[fake_data$j]
     fake_data$Y1 <- with(fake_data,mean(Y0) + treatment_effect + (Y0 - mean(Y0)) * (2 / 3))
     fake_data$Z  <- ifelse(fake_data$j %in% sample(1:J,J / 2) == TRUE, 1, 0)
     fake_data$Y  <- with(fake_data, Z * Y1 + (1 - Z) * Y0)
     return(fake_data)
}

set.seed(12345)
pretend_data <- make_clustered_data(J = 10,n = 100,treatment_effect = .25,ICC = .1)
```

Because we have created the data ourselves, we can calculate the true standard
error of our estimator. We firstly generate the true sampling distribution by
simulating every possible permutation of the treatment and calculating the
estimate each time. The standard deviation of this distribution is the standard
error of the estimator.

<div class="hidecode" onclick="doclick(this);">[Click to show code]</div>
```{r,cache=T}
# Define the number of clusters
J <- length(unique(pretend_data$j))
# Generate all possible ways of combining clusters into a treatment group
all_treatment_groups <- with(pretend_data,combn(x = 1:J,m = J/2))
# Create a function for estimating the effect
clustered_ATE <- function(j,Y1,Y0,treated_clusters) {
     Z_sim    <- (j %in% treated_clusters)*1
     Y        <- Z_sim * Y1 + (1 - Z_sim) * Y0
     estimate <- mean(Y[Z_sim == 1]) - mean(Y[Z_sim == 0])
     return(estimate)
}

set.seed(12345)
# Apply the function through all possible treatment assignments
cluster_results <- apply(
  X = all_treatment_groups, MARGIN = 2,
                         FUN = clustered_ATE,
                         j  = pretend_data$j,Y1 = pretend_data$Y1,
  Y0 = pretend_data$Y0
)

true_SE <- sd(cluster_results)

true_SE
```

This gives a standard error of `r round(true_SE,2)`. We can compare the true
standard error to two other kinds of standard error commonly employed. The first
ignores clustering and assumes that treatment effect estimates are identically
and independently distributed according to a normal distribution. We will refer
to this as the I.I.D. standard error. To take clustering into account, we can
use the following formula for the standard error:

$$\text{Var}_\text{clustered}(\hat{\tau})=\frac{\sigma^2}{\sum_{j=1}^J \sum_{i=1}^{n_j} (Z_{ij}-\bar{Z})^2} (1-(n-1)\rho)$$

where $\sigma^2=\sum_{j=1}^J \sum_{i=1}^{n_j} (Y_{ij} - \bar{Y}_{ij})^2$ (following @arceneaux2009modeling ).
This adjustment to the IID standard error
is commonly known as the “Robust Clustered Standard Error” or RCSE.


<div class="hidecode" onclick="doclick(this);">[Click to show code]</div>
```{r,cache=T, error=FALSE, message=FALSE, warning=FALSE}

ATE_estimate <- lm(Y ~ Z,data = pretend_data)

IID_SE <- function(model) {
     return(sqrt(diag(vcov(model)))[["Z"]])
}

RCSE <- function(model, cluster,return_cov = FALSE){
  require(sandwich)
  require(lmtest)
  M <- length(unique(cluster))
  N <- length(cluster)
  K <- model$rank
  dfc <- (M/(M - 1)) * ((N - 1)/(N - K))
  uj <- apply(estfun(model), 2, function(x) tapply(x, cluster, sum))
  rcse.cov <- dfc * sandwich(model, meat = crossprod(uj)/N)
  rcse.se <- as.matrix(coeftest(model, rcse.cov))
  if(return_cov){
    return(rcse.cov)
  }else{
    return(rcse.se)
  }
}

IID_SE_estimate <- IID_SE(model = ATE_estimate)

RCSE_estimate   <- RCSE(model = ATE_estimate,cluster = pretend_data$j)

knitr::kable(round(
  data.frame(
     true_SE         = true_SE,
     IID_SE_estimate = IID_SE_estimate,
    RCSE_estimate   = RCSE_estimate["Z", "Std. Error"]
  ),
  2
))
```

When we ignore the clustered-assignment, the standard error is too small: we are
over-confident about the amount of information provided to us by the experiment.
The RCSE is slightly more conservative than the true standard error in this
instance, but is very close. The discrepancy is likely because the RCSE is not a
good approximation of the true standard error when the number of clusters is as
small as it is here. To illustrate the point further, we can compare a
simulation of the true standard error generated through random permutations of
the treatment to the IID and RC standard errors.


<div class="hidecode" onclick="doclick(this);">[Click to show code]</div>
```{r,cache=T}

compare_SEs <- function(data) {
     simulated_SE <- sd(replicate(
          5000,
          clustered_ATE(
               j = data$j,
               Y1 = data$Y1,
               Y0 = data$Y0,
               treated_clusters = sample(unique(data$j),length(unique(data$j))/2)
    )
  ))
     ATE_estimate <- lm(Y ~ Z,data)
     IID_SE_estimate <- IID_SE(model = ATE_estimate)
     RCSE_estimate <- RCSE(model = ATE_estimate,cluster = data$j)["Z", "Std. Error"]
     return(round(c(
          simulated_SE = simulated_SE,
          IID_SE = IID_SE_estimate,
          RCSE = RCSE_estimate
     ),3))
}

J_4_clusters    <- make_clustered_data(J = 4)
J_10_clusters   <- make_clustered_data(J = 10)
J_30_clusters   <- make_clustered_data(J = 30)
J_100_clusters  <- make_clustered_data(J = 100)
J_1000_clusters <- make_clustered_data(J = 1000)

set.seed(12345)

knitr::kable(rbind(
  c(J = 4,compare_SEs(J_4_clusters)),
  c(J = 30,compare_SEs(J_30_clusters)),
  c(J = 100,compare_SEs(J_100_clusters)),
  c(J = 1000,compare_SEs(J_1000_clusters))
  ))
```

As these simple examples illustrate, the clustered estimate of the standard
error (RCSE) gets closer to the truth (the simulated standard error) as the
number of clusters increases. Meanwhile, the standard error ignoring clustering
(assuming IID) tends to be smaller than either of the other standard errors. The
smaller the estimate of the standard error is, the more precise the estimates
seem to us, and the more likely we will find results that appear 'statistically
significant'. This is problematic: in this case, the IID standard error is leads
us to be over confident in our results because it ignores intra-cluster
correlation, the extent to which differences between units can be attributed to
the cluster they are a member of. If we estimate standard errors using
techniques that understate our uncertainty, we are more likely to falsely reject
null hypotheses when we should not.

Another way to approach the problems that clustering introduces into the
calculation of standard errors is to analyze the data at the level of the
cluster. In this approach, we take averages or sums of the outcomes within the
clusters, and then treat the study as though it only took place at the level of
the cluster. @hansen2008covariate show that we can characterize the distribution
of the difference of means using what we know about the distribution of the
*sum* of the outcome in the treatment group, which varies from one assignment of
treatment to another.


<div class="hidecode" onclick="doclick(this);">[Click to show code]</div>
```{r,cache=T}
# Aggregate the unit-level data to the cluster level
# Sum outcome to cluster level
Yj <- tapply(pretend_data$Y,pretend_data$j,sum)
# Aggregate assignment indicator to cluster level
Zj <- tapply(pretend_data$Z,pretend_data$j,unique)
# Calculate unique cluster size
n_j <- unique(as.vector(table(pretend_data$j)))
# Calculate total sample size (our units are now clusters)
N <- length(Zj)
# Generate cluster id
j <- 1:N
# Calculate number of clusters treated
J_treated <- sum(Zj)

# Make a function for the cluster-level difference in means estimator (See Hansen & Bowers 2008)
cluster_difference <- function(Yj,Zj,n_j,J_treated,N){
     ones <- rep(1, length(Zj))
     ATE_estimate <- crossprod(Zj,Yj)*(N/(n_j*J_treated*(N-J_treated))) -
          crossprod(ones,Yj)/(n_j*(N-J_treated))
     return(ATE_estimate)
}

# Given equal sized clusters and no blocking, this is identical to the
# unit-level difference in means

ATEs <- colMeans(data.frame(
  cluster_level_ATE =
               cluster_difference(Yj,Zj,n_j,J_treated,N),
          unit_level_ATE =
    with(pretend_data, mean(Y[Z == 1]) - mean(Y[Z == 0]))
))

knitr::kable(data.frame(ATEs),align = "c")
```

In order to characterize uncertainty about the cluster-level ATE, we can
exploit the fact that the only random element of the estimator is now the
cross-product between the cluster-level assignment vector and the cluster-level
outcome, $\mathbf{Z}^\top\mathbf{Y}$, scaled by some constant.
We can estimate the variance of this random component through permutation
of the assignment vector or through an approximation of the variance, assuming
that the sampling distribution follows a normal distribution.

<div class="hidecode" onclick="doclick(this);">[Click to show code]</div>
```{r,cache=T}
# Approximating variance using normality assumptions
normal_sampling_variance <-
     (N/(n_j*J_treated*(N-J_treated)))*(var(Yj)/n_j)
# Approximating variance using permutations
set.seed(12345)
sampling_distribution <- replicate(10000,cluster_difference(Yj,sample(Zj),n_j,J_treated,N))

ses <- data.frame(
  sampling_variance = c(sqrt(normal_sampling_variance), sd(sampling_distribution)),
  p_values = c(
    2 * (1 - pnorm(abs(ATEs[1]) / sqrt(normal_sampling_variance), mean = 0)),
                               2*min(mean(sampling_distribution>=ATEs[1]),mean(sampling_distribution<=ATEs[1]))
  )
)

rownames(ses) <- c("Assuming Normality","Permutations")

knitr::kable(ses)
```

This cluster-level approach has the advantage of correctly characterizing
uncertainty about effects when randomization is clustered, without having to use
the RCSE standard errors for the unit-level estimates, which are overly
permissive for small N. Indeed, the false positive rate of tests based on RCSE
standard errors tend to be incorrect when the number of clusters is small,
leading to over-confidence. As we shall see below, however, when the number of
clusters is very small ($J=4$) the cluster-level approach is overly
conservative, rejecting the null with a probability of 1. Another drawback of
the cluster-level approach is that it does not allow for the estimation of
unit-level quantities of interest, such as heterogeneous treatment effects.

# 4 Why clustering can matter II: different cluster sizes and bias

When clusters are of different sizes, this can pose a unique class of problems
related to the estimation of the treatment effect. Especially when the size of
the cluster is in some way related to the potential outcomes of the units within
it many conventional estimators of the sample average treatment effect (SATE)
can be biased.

To fix ideas, imagine an intervention targeted at firms of different sizes,
which seeks to increase worker productivity. Due to economies of scale, the
productivity of employees in big firms is increased much more proportional to
that of employees in smaller firms. Imagine that the experiment includes 20
firms ranging in size from one-person entrepreneurs to large outfits with over
500 employees. Half of the firms are assigned to the treatment, and the other
half are assigned to control. Outcomes are defined at the employee level.

<div class="hidecode" onclick="doclick(this);">[Click to show code]</div>
```{r,cache=T}
set.seed(1000)
# Number of firms
J <- 20

# Employees per firm
n_j <- rep(2^(0:(J/2-1)),rep(2,J/2))

# Total number of employees
N <- sum(n_j)
# 2046

# Unique employee (unit) ID
i <- 1:N

# Unique firm (cluster) ID
j <- rep(1:length(n_j),n_j)

# Firm specific treatment effects
cluster_ATE <- n_j^2/10000

# Untreated productivity
Y0 <- rnorm(N)

# Treated productivity
Y1 <- Y0 + cluster_ATE[j]

# True sample average treatment effect
(true_SATE <- mean(Y1-Y0))
```

<div class="hidecode" onclick="doclick(this);">[Click to show code]</div>
```{r,cache=T}
# Correlation between firm size and effect size
cor(n_j,cluster_ATE)
```

As we see, there is high correlation in the treatment effect and cluster size.
Now let us simulate 1000 analyses of this experiment, permuting the treatment
assignment vector each time, and taking the unweighted difference in means as
an estimate of the sample average treatment effect.


<div class="hidecode" onclick="doclick(this);">[Click to show code]</div>
```{r,cache=T}
set.seed(1234)
# Unweighted SATE
SATE_estimate_no_weights <- NA
for(i in 1:1000){
     # Clustered random assignment of half of the firms
     Z <- (j %in% sample(1:J,J/2))*1
     # Reveal outcomes
     Y <- Z*Y1 + (1-Z)*Y0
     # Estimate SATE
     SATE_estimate_no_weights[i] <- mean(Y[Z==1])-mean(Y[Z==0])
     }

# Generate histogram of estimated effects
hist(SATE_estimate_no_weights,xlim = c(true_SATE-2,true_SATE+2),breaks = 100)
# Add the expected estimate of the SATE using this estimator
abline(v=mean(SATE_estimate_no_weights),col="blue")
# And add the true SATE
abline(v=true_SATE,col="red")
```

The histogram shows the sampling distribution of the estimator, with the true
SATE in red and the unweighted estimate thereof in blue. The estimator is
biased: in expectation, we do not recover the true SATE, instead underestimating
it. Intuitively, one might correctly expect that the problem is related to the
relative weight of the clusters in the calculation of the treatment effect.
However, in this situation, taking the difference in the weighted average of the
outcome among treated and control clusters is not enough to provide an unbiased
estimator (see the code below).


<div class="hidecode" onclick="doclick(this);">[Click to show code]</div>
```{r,cache=T}
set.seed(1234)
# Weighted cluster-averages
SATE_estimate_weighted <- NA
for(i in 1:1000){
     # Define the clusters put into treatment
     treated_clusters <- sample(1:J,J/2,replace = F)
     # Generate unit-level assignment vector
     Z <- (j %in% treated_clusters)*1
     # Reveal outcomes
     Y <- Z*Y1 + (1-Z)*Y0
     # Calculate the cluster weights
     treated_weights <- n_j[1:J%in%treated_clusters]/sum(n_j[1:J%in%treated_clusters])
     control_weights <- n_j[!1:J%in%treated_clusters]/sum(n_j[!1:J%in%treated_clusters])
     # Calculate the means of each cluster
     treated_means <- tapply(Y,j,mean)[1:J%in%treated_clusters]
     control_means <- tapply(Y,j,mean)[!1:J%in%treated_clusters]
     # Calculate the cluster-weighted estimate of the SATE
     SATE_estimate_weighted[i] <-
          weighted.mean(treated_means,treated_weights) -
          weighted.mean(control_means,control_weights)
}

# Generate histogram of estimated effects
hist(SATE_estimate_weighted,xlim = c(true_SATE-2,true_SATE+2),breaks = 100)
# Add the expected estimate of the unweighted SATE
abline(v=mean(SATE_estimate_no_weights),col="blue")
# Add the expected estimate of the weighted SATE
abline(v=mean(SATE_estimate_weighted),col="green")
# And add the true SATE
abline(v=true_SATE,col="red")
```

The histogram shows the sampling distribution of the weighted estimator, with the
true SATE in red and the unweighted estimate in blue, and the weighted estimate in green.
In expectation, the weighted version of the estimator in fact gives the same
estimate of the SATE as the non-weighted version. What is the nature of the bias?

Instead of assigning treatment to half of the clusters
and comparing outcomes at the level of the 'treatment' and 'control'
groups, imagine that we paired each cluster with one other
cluster, and assigned one to treatment within each pair. Our estimate of the treatment
effect is then the aggregate of the pair-level estimates.
This is analogous to the complete random assignment procedure employed
above, in which $J/2$ firms were assigned to treatment. Now, we will instead
refer to the $k$'th of the $m$ pairs, where $2m = J$.

Given this setup, @imai2009essential give the following formal definition of the
bias in the cluster-weighted difference-in-means estimator

$$\frac{1}{n}
\sum^m_{k = 1}
\sum^2_{l = 1}
\left[
\left(
\frac{n_{1k} + n_{2k}}{{2} - n_{lk}}
\right)
\times
\sum^{n_{lk}}_{i = 1}
\frac{Y_{ilk}(1) - Y_{ilk}(0)}{n_{lk}}
\right],$$

where $l = 1,2$ indexes the clusters within each pair. Thus,
$n_{1k}$ refers to the number of units in the first of the $k$'th
pair of clusters.

This expression indicates that bias from unequal cluster sizes arises if and
only if two conditions are met. Firstly, the sizes of at least one pair of
clusters must be unequal: when $n_{1k}=n_{2k}$ for all $k$, the bias term is
reduced to 0. Secondly, the weighted effect sizes of at least one pair of
clusters must be unequal: when $\sum_{i =
1}^{n_{1k}}(Y_{i1k}(1)-Y_{i1k}(0))/n_{1k} = \sum_{i =
1}^{n_{2k}}(Y_{i2k}(1)-Y_{i2k}(0))/n_{2k}$ for all $k$, the bias is also reduced
to 0.

# 5 What to do about different cluster sizes

Two other approaches work to alleviate this problem of bias from the correlation
of cluster size and treatment effect: (1) condition on cluster size directly and
(2) use the Horvitz-Thompson estimator. This [DeclareDesign Blog
Post](https://declaredesign.org/blog/bias-cluster-randomized-trials.html)
discusses both options. Here we demonstrate a pairing approach to conditioning
on cluster-size given the example data that we generated above.

As the above expression suggests, in order to reduce the bias from unequal
cluster sizes to almost 0, it is sufficient to put clusters into pairs that
either are of equal size or  have almost identical potential outcomes.

We demonstrate this approach below using the same data as we examined in the
example of a hypothetical firm-randomized employee productivity experiment.

<style>
div.hidecode + pre {display: none}
</style>
<script>
doclick=function(e){
e.nextSibling.nextSibling.style.display="block";
}
</script>

<div class="hidecode" onclick="doclick(this);">[Click to show code]</div>
```{r,cache=T}
set.seed(1234)
# Make a function that matches pairs based on size
pair_sizes <- function(j,n_j){
     # Find all of the unique sizes
     unique_sizes <- unique(n_j)
     # Find the number of unique sizes
     N_unique_sizes <- length(unique_sizes)
     # Generate a list of candidates for pairing at each cluster size
  possible_pairs <- lapply(unique_sizes, function(size) {
    which(n_j == size)
  })
     # Find the number of all possible pairs (m)
     m_pairs <- length(unlist(possible_pairs))/2
     # Generate a vector with unique pair-level identifiers
     pair_indicator <- rep(1:m_pairs,rep(2,m_pairs))
     # Randomly assign units of the same cluster size into pairs
     pair_assignment <-
          unlist(lapply(
               possible_pairs,
               function(pair_list){
        sample(pair_indicator[unlist(possible_pairs) %in% pair_list])
      }
    ))
     # Generate a vector indicating the k'th pair for each i unit
     pair_assignment <- pair_assignment[match(x = j,table = unlist(possible_pairs))]
     return(pair_assignment)
}

pair_indicator <- pair_sizes(j = j , n_j = n_j)

SATE_estimate_paired <- NA
for(i in 1:1000){
     # Now loop through the vector of paired assignments
     pair_ATEs <- sapply(unique(pair_indicator),function(pair){
          # For each pair, randomly assign one to treatment
          Z <- j[pair_indicator==pair] %in% sample(j[pair_indicator==pair],1)*1
          # Reveal the potential outcomes of the pair
          Y <- Z*Y1[pair_indicator==pair] + (1-Z)*Y0[pair_indicator==pair]
          clust_weight <- length(j[pair_indicator==pair])/N
          clust_ATE <- mean(Y[Z==1])-mean(Y[Z==0])
          return(c(weight = clust_weight, ATE = clust_ATE))
     })

     SATE_estimate_paired[i] <- weighted.mean(x = pair_ATEs["ATE",],w = pair_ATEs["weight",])
}

# Generate histogram of estimated effects
hist(SATE_estimate_paired,xlim = c(true_SATE-2,true_SATE+2),breaks = 100)
# Add the expected estimate of the paired SATE
abline(v=mean(SATE_estimate_paired),col="purple")
# And add the true SATE
abline(v=true_SATE,col="red")
```

<style>
div.hidecode + pre {display: none}
</style>
<script>
doclick=function(e){
e.nextSibling.nextSibling.style.display="block";
}
</script>

<div class="hidecode" onclick="doclick(this);">[Click to show code]</div>
```{r,cache=T}
# The paired estimator is much closer to the true SATE
kable(round(data.frame(
  true_SATE = true_SATE,
  paired_SATE = mean(SATE_estimate_paired),
  weighted_SATE = mean(SATE_estimate_weighted),
  unweighted_SATE = mean(SATE_estimate_no_weights)
),2))
```

In spite of unequal cluster sizes, the bias is completely eliminated
by this technique: in expectation, the paired estimator recovers
the true Sample Average Treatment Effect, whereas the cluster-weighted
and non-weighted difference-in-means estimators are biased.

Note also that the variance in the sampling distribution is much lower
for the pair-matched estimator, giving rise to much more precise
estimates. Thus, pair-matching not only promises to reduce bias,
but can also greatly mitigate the problem of information reduction
that clustering induces.

Such pre-randomization pair matching, however, does impose some
constraints on the study, some of which may be difficult to meet
in practice. For example, it may be difficult or even impossible
to find perfectly-matched pairs for every cluster size, especially
when there are multiple treatments (such that, instead of pairs,
treatment is randomized over triplets or quadruplets). In such
cases, researchers may adopt several other solutions, such as
creating pairs by matching on observed covariates prior to the
randomization, whereby, for example, the within-pair similarity of
observed covariates is maximized. @imai2009essential
recommend a mixture model for post-randomization pair-matched
estimation, and spell out some of the assumptions that must be
made for such estimates to be valid. And this  [DeclareDesign Blog
Post](https://declaredesign.org/blog/bias-cluster-randomized-trials.html)
demonstrates conditioning on cluster-size without strict pairing.


# 6 Why clustering can matter III: within-cluster spillovers

In many, or most experiments, we would like to estimate the average causal
effect of the treatment within a population or a sample.
Denoting $Y_{z_i}$ the outcome $Y$ of unit $i$ when assigned to the treatment
status $z_i \in \{1,0\}$, we can define
this quantity -- the ATE (Average Treatment Effect) --
as the expected value of the difference
between the sample when assigned to treatment, $Y_1$ and the sample when
assigned to control $Y_0$: $E[Y_1 - Y_0]$.

However, it may be the case that a unit's outcome depends on the treatment
status $z_j$ of another unit, $j$, inside the same cluster. In that case,
we denote potential outcomes
$Y_{z_j,z_i} \in \{ Y_{00}, Y_{10}, Y_{01}, Y_{11} \}$,
where an untreated unit with an untreated cluster neighbor is defined
as $Y_{00}$, an untreated unit with a treated cluster neighbor as $Y_{10}$,
a treated unit with an untreated cluster neighbor as $Y_{01}$, and
a treated unit with a treated cluster neighbor as $Y_{11}$.
When we conduct a
cluster-randomized experiment, we typically assume that a unit's outcome
is not a function of the treatment status of the units with whom it shares
a cluster, or formally $Y_{01}=Y_{11}=Y_1$ and $Y_{10}=Y_{00}=Y_0$.
Yet, for all sorts of reasons this may not be the case: depending on whom
someone finds themselves in the same cluster with, and whether or not that
cluster is assigned to treatment, their outcomes may be very different.

Consider an experiment in which five pairs of students living in dorms are
randomly assigned to either receive or not receive a food subsidy, and their
stated well-being is the outcome of interest. Let us assume that four students
are vegetarian (V) and six are meat-eaters (M). When a VV, MM, or VM pair is
assigned to control, they do not receive the subsidy and their well-being is
unaffected. However, when assigned to treatment, VM pairs quarrel and this
reduces their well-being, whereas VV and MM pairs do not fight and
are affected only by the treatment. Let us denote $x_k \in \{0,1\}$ an
indicator for whether the pair is mismatched, where the unit's outcome
is denoted $Y_{z_j,z_j,x_k}$. This implies that
$Y_{110} = Y_1$ and $Y_{000} = Y_{001} = Y_0$, whereas
$Y_{111} \neq Y_1$.
To understand how this matters, let us simulate such an experiment.

<style>
div.hidecode + pre {display: none}
</style>
<script>
doclick=function(e){
e.nextSibling.nextSibling.style.display="block";
}
</script>

<div class="hidecode" onclick="doclick(this);">[Click to show code]</div>
```{r,cache = T}
# Create experimental data
N <- 10
types <- c(rep("V",.4*N),rep("M",.6*N))
ID <- 1:length(types)
baseline <- rnorm(length(ID))

# The true treatment effect is 5
true_ATE <- 5
# If a pair is mismatched (VM, MV), they get a spillover of -10
spillover_or_not <- function(type_i,type_j){
  ifelse(type_i==type_j,yes = 0,no = -10)
}

# A function for forming pairs
form_pairs <- function(ID,types){
  N <- length(ID)
  k <- rep(1:(N/2),2)
  pair_place <- rep(c(1,2),c(N/2,N/2))
  ID_draw <- sample(ID)
  type_i <- types[ID_draw]
  pair_1 <- type_i[pair_place==1]
  pair_2 <- type_i[pair_place==2]
  ID_j_1 <- ID_draw[pair_place==1]
  ID_j_2 <- ID_draw[pair_place==2]
  type_j <- c(pair_2,pair_1)
  j <- c(ID_j_2,ID_j_1)
  return(data.frame(i = ID_draw,j = j,k = k, type_i = type_i, type_j = type_j))
}


# A function for assigning treatment and revealing the outcome
assign_reveal_est <- function(k,i,effect,spillover){
  Z <- (k %in% sample(k,N/2))*1
  Y <- baseline[i] + Z*effect + Z*spillover
  mean(Y[Z==1])-mean(Y[Z==0])
}

# A function for simulating the experiment
simulate_exp <- function(){
  data <- form_pairs(ID,types)
  spillover <- spillover_or_not(data$type_i,data$type_j)
  estimate <- assign_reveal_est(k = data$k,effect = true_ATE,spillover = spillover,i = data$i)
  return(estimate)
}

# Estimate the effects one thousand times
est_effects <- replicate(n = 1000,expr = simulate_exp())

# Plot the estimates as bars, the expected ATE in blue, and the true ATE in red
hist(est_effects,breaks = 100,xlim = c(-7,7))
abline(v = true_ATE,col = "red")
abline(v = mean(est_effects,na.rm = T),col = "blue")
```

As the plot above shows, this is a biased estimator of the true
individual level treatment effect, $Y_{01} - Y_{00}$.
In expectation, we estimate an effect close to 0, obtaining
very negative effects in almost half of the simulations of this
experiment.
The key point here is that the estimand is changed: rather than the
ATE, we obtain a combination
of the true treatment effect among those who are matched (do not
experience spillovers) $E[Y_{110}-Y_{00x_k}]$, and the combined treatment
and spillover effect for those who are unmatched $E[Y_{111}-Y_{00x_k}]$.
Crucially, however, we cannot identify the impact of the spillover,
$E[Y_{101}-Y_{00x_k}]$, independently of the direct effect. This is because the
randomization is clustered: it is not possible to observe
$Y_{101}$ in a cluster-randomized scheme, because all units within a
cluster are always treated.
Generally speaking, this issue is true of any cluster-randomized study:
in order to make the claim that we identify the individual-level effect of the
treatment, we must assume that $Y_{11}=Y_{1}$ and $Y_{00}=Y_{0}$.

# 7 What to do about within-cluster spillovers

If there are strong reasons to believe that intra-cluster spillovers occur,
then researchers can take different approaches depending on the manner in
which clusters are formed.
In some studies researchers must themselves sort units into groups for the
purposes of experimentation: for example, in a study involving
a vocational programme, the researcher may be able to decide who is
recruited into which class. In such cases, if the researcher can make
plausible assumptions about spillovers, then the individual-level
treatment effect may be recoverable.

Consider a researcher conducting the previous study above who correctly
assumed that spillovers would occur between mismatched pairs. In this case,
the researcher can recover the true individual treatment effect
by forming clusters that are not
susceptible to spillover.

<style>
div.hidecode + pre {display: none}
</style>
<script>
doclick=function(e){
e.nextSibling.nextSibling.style.display="block";
}
</script>

<div class="hidecode" onclick="doclick(this);">[Click to show code]</div>
```{r,cache = T}
form_matched_pairs <- function(ID,types){
  pair_list <- lapply(unique(types),function(type){
    ID <- ID[types == type]
    types <- types[types == type]
    draw_ID <- 1:length(types)
    matched_pairs <- form_pairs(ID = draw_ID,types = types)
    matched_pairs$i <- ID[matched_pairs$i]
    matched_pairs$j <- ID[matched_pairs$j]
    matched_pairs$k <- paste0(type,"_",matched_pairs$k)
    return(matched_pairs)
  })
  data <- rbind(pair_list[[1]],pair_list[[2]])
  return(data)
}

simulate_matched_exp <- function(){
  data <- form_matched_pairs(ID,types)
  spillover <- spillover_or_not(data$type_i,data$type_j)
  estimate <- assign_reveal_est(k = data$k,effect = true_ATE,spillover = spillover,i = data$i)
  return(estimate)
}

# Estimate the effects one thousand times
est_matched_effects <- replicate(n = 1000,expr = simulate_matched_exp())

hist(est_matched_effects,breaks = 100,xlim = c(-7,7))
abline(v = true_ATE,col = "red")
abline(v = mean(est_matched_effects,na.rm = T),col = "blue")
```


In the case that researchers are not able to control how clusters are formed,
they can still investigate cluster-level heterogeneity in treatment effects
as a way of understanding possible spillovers. However, in both cases,
assumptions must be made about the nature of spillovers. Strictly speaking,
these cannot be causally identified due to the unobservability of the
outcomes $Y_{01}$ and $Y_{10}$. Ultimately, one would need to combine
clustered and non-clustered randomization schemes in order to
estimate the effects of intra-cluster spillover, $Y_{11} - Y_{01}$ and
 $Y_{01} - Y_{00}$. Therefore, in the interests of interpreting results
 correctly,
 researchers should be careful
 when defining their estimand to take account of the potential for
 intra-cluster spillover.

# 8 Performance of design- vs. model-based analyses of clustered studies

In our discussion of information loss, we assessed
approaches that require (1) that treatment was randomized as planned and
(2) that the treatment assigned to one unit did not change the potential outcomes for any other unit.
In cases where these assumptions may be violated, it is sometimes simpler to specify statistical models that attempt to describe the features of complex designs. Even if we do not believe the models as scientific descriptions of a known process, this can be a more informative and flexible way of analyzing an experiment than to derive complex new expressions for design-based estimators.

In model-based approaches, the sampling distribution of an estimator is approximated
using probability distributions to characterize our uncertainty about
unknown quantities, such as the true treatment effect or the true mean of the outcome at the
cluster level.
Such approaches are referred to as 'model-based', because they depict causal relationships as arising from interrelated probability distributions. Often, such approaches use
'multilevel models', in which unknown parameters --- such as differences between clusters --- are themselves understood as arising from probability distributions.
Thus, for example, there might be a model for individual-level outcomes,
whose intercept and/or coefficients vary from one cluster to another.
In this manner, it is possible to model the 'effect of being a unit in cluster A',
separately from the estimation of the treatment effect.
The advantage of such approaches is that they allow for 'partial pooling' of the
variance in the population and the variance among clusters. When a given cluster is
poorly estimated, it contributes less weight to the estimation, and vice versa.
Such models therefore often work well in situations where there is very little data
in some clusters: through the specification of a Bayesian posterior distribution,
they are able to leverage information from all parts of the study.
The trade-off is that such assumption-heavy models are only correct to the extent that
the assumptions underlying them are correct.

Here we show that the estimated effect is the same whether we use a simple difference of means (via OLS) or a multilevel model in our very simplified cluster randomized trial setup.

<style>
div.hidecode + pre {display: none}
</style>
<script>
doclick=function(e){
e.nextSibling.nextSibling.style.display="block";
}
</script>

<div class="hidecode" onclick="doclick(this);">[Click to show code]</div>
```{r,cache = T, error=FALSE, warning=FALSE, message=FALSE}
library(lme4)

simple_OLS <- lm(Y ~ Z,data = J_30_clusters)
multilevel <- lmer(Y ~ Z + (1 | j),
                   data = J_30_clusters,
  control = lmerControl(optimizer = "bobyqa"),
  REML = TRUE
)

kable(round(data.frame(
  OLS=coef(simple_OLS)["Z"],Multilevel=fixef(multilevel)["Z"]
  ),3))
```

The confidence intervals differ even though the estimates are the same --- and
there is more than one way to calculate confidence intervals and hypothesis
tests for multilevel models. The software in R [@bates_et_al_2014lme4] includes
three methods by default and @gelman2006data recommend MCMC sampling from the
implied posterior. Here we focus on the Wald method only because it is the
fastest to compute.

<style>
div.hidecode + pre {display: none}
</style>
<script>
doclick=function(e){
e.nextSibling.nextSibling.style.display="block";
}
</script>

<div class="hidecode" onclick="doclick(this);">[Click to show code]</div>
```{r,cache = T, error=FALSE, message=FALSE, warning=FALSE}
# This function calculates confidence intervals for linear models
# with custom variance-covariance matrices
confint_HC<-function (coefficients, df, level = 0.95, vcov_mat, ...) {
  a <- (1 - level)/2
  a <- c(a, 1 - a)
  fac <- qt(a, df)
  ses <- sqrt(diag(vcov_mat))
  coefficients + ses %o% fac
}

simple_OLS_CI <-
  confint_HC(
    coefficients = coef(simple_OLS),
    vcov_mat = RCSE(
      model = simple_OLS,
                             cluster = J_30_clusters$j,
      return_cov = TRUE
    ),
    df = simple_OLS$df.residual
  )["Z", ]

multi_wald_CI <- lme4::confint.merMod(
  multilevel,
  parm = "Z", method = "Wald"
)["Z", ]
multi_profile_CI <- lme4::confint.merMod(
  multilevel,
  parm = 4, method = "profile"
)["Z", ]


knitr::kable(round(rbind(
  Design_Based_CI = simple_OLS_CI,
  Model_Based_Wald_CI = multi_wald_CI,
  Model_Based_Profile_CI = multi_profile_CI
),3))
```

We can calculate an estimate of the ICC directly from the model quantities (the variance of the Normal prior that represents the cluster-to-cluster differences in the intercept over the total variance of the Normal posterior).

<style>
div.hidecode + pre {display: none}
</style>
<script>
doclick=function(e){
e.nextSibling.nextSibling.style.display="block";
}
</script>

<div class="hidecode" onclick="doclick(this);">[Click to show code]</div>
```{r,cache = T}
VC <- as.data.frame(lme4:::VarCorr.merMod(multilevel))
round(c(ICC = VC$vcov[1] / sum(VC$vcov)),2)
```

In order to assess the performance of this model-based approach, as opposed to the
robust-clustered standard-error (RCSE) and cluster-aggregated approaches outlined
above, we can check how often the different approaches falsely reject the
sharp null hypothesis of no effects for any unit, when we know that this null is
true.

To do so, we write a function that firstly breaks the relationship between the
treatment assignment and the outcome by randomly shuffling the assignment, and then
tests whether 0 is in the 95% confidence interval for each of the three approaches,
as it should be.
Recall that, valid tests would have error rates within 2 simulation standard errors of .95 - this would mean that a correct null hypothesis would be rejected no more than 5% of the time.

<style>
div.hidecode + pre {display: none}
</style>
<script>
doclick=function(e){
e.nextSibling.nextSibling.style.display="block";
}
</script>

<div class="hidecode" onclick="doclick(this);">[Click to show code]</div>
```{r,cache = T, message=FALSE}
# Make a function for checking whether 0 is in the confidence interval of
# the RCSE, cluster-aggregated, and multilevel estimation approaches

sim_0_ate <- function(J,Y) {
  #   Make the true relationship between treatment and outcomes equal zero by
  #   shuffling Z but not revealing new potential outcomes
  z.sim <- sample(1:max(J), max(J) / 2)
  Z_new <- ifelse(J %in% z.sim == TRUE, 1, 0)

  # Estimate using the linear model for RCSE
  linear_fit <- lm(Y ~ Z_new)
  linear_RCSE <- RCSE(
    model = linear_fit,
                  cluster = J,
    return_cov = TRUE
  )
  linear_CI <- confint_HC(
    coefficients = coef(linear_fit),
                      vcov_mat = linear_RCSE,
    df = linear_fit$df.residual
  )["Z_new", ]
  # Check if the confidence interval bounds 0
  zero_in_CI_RCSE <- (0 >= linear_CI[1]) & (0 <= linear_CI[2])

  # Estimate using cluster-aggregated approach (Hansen and Bowers 2008)
  Yj <- tapply(Y, J, sum)
  Zj <- tapply(Z_new, J, mean)
  m0 <- unique(table(J))
  n  <- length(Zj)
  nt <- sum(Zj)
  # Do Hansen and Bowers 2008 based test for difference of means
  # with cluster-level assignment (assuming same size clusters)
  ones <- rep(1, length(Yj))
  dp   <- crossprod(Zj,Yj) * (n / (m0 * nt * (n - nt))) -
    crossprod(ones,Yj) / (m0 * (n - nt))
  obs_ATE <- dp[1,1]
  # Two tailed p-value for the test of the null of no effects
  Vdp <- (n / (m0 * nt * (n - nt))) * (var(Yj) / m0)
  HB_pval <- 2 * (1 - pnorm(abs(obs_ATE) / sqrt(Vdp)))
  # Check if the p-value is greater than .05
  zero_not_rej_HB <- HB_pval >= .05

  # Estimate using a multilevel model
  multilevel_fit <- lmer(Y ~ Z_new + (1 | J),
    control = lmerControl(optimizer = "bobyqa"),
    REML = FALSE
  )

  multilevel_CI <- lme4:::confint.merMod(
    multilevel_fit,
    parm = "Z_new", method = "Wald"
  )
  # Check if the confidence interval bounds 0
  zero_in_CI_multilevel <- (0 >= multilevel_CI[1]) & (0 <= multilevel_CI[2])

  return(
    c(
      ATE = fixef(multilevel_fit)["Z_new"],
      zero_in_CI_RCSE = zero_in_CI_RCSE,
      zero_not_rej_HB = zero_not_rej_HB,
      zero_in_CI_multilevel = zero_in_CI_multilevel
    )
  )
}

# Now simulate each of the estimates 1000 times

J_4_comparison <- replicate(1000, sim_0_ate(J = J_4_clusters$j, Y = J_4_clusters$Y))
J_4_error_rates <- apply(J_4_comparison,1,mean)
J_4_error_rates[-1] <- 1-J_4_error_rates[-1]

J_10_comparison <- replicate(1000, sim_0_ate(J = J_10_clusters$j, Y = J_10_clusters$Y))
J_10_error_rates <- apply(J_10_comparison,1,mean)
J_10_error_rates[-1] <- 1-J_10_error_rates[-1]

J_30_comparison <- replicate(1000, sim_0_ate(J = J_30_clusters$j, Y = J_30_clusters$Y))
J_30_error_rates <- apply(J_30_comparison,1,mean)
J_30_error_rates[-1] <- 1-J_30_error_rates[-1]

J_100_comparison <- replicate(1000, sim_0_ate(J = J_100_clusters$j, Y = J_100_clusters$Y))
J_100_error_rates <- apply(J_100_comparison,1,mean)
J_100_error_rates[-1] <- 1-J_100_error_rates[-1]

error_comparison <- data.frame(round(rbind(
  J_4_error_rates,
  J_10_error_rates,
  J_30_error_rates,
  J_100_error_rates
),3))

colnames(error_comparison) <- c(
  "Estimated ATE",
                                "OLS + RCSE",
                                "Cluster-Level",
  "Multi-Level"
)

kable(error_comparison,align = "c")
```

In our simple setup, the individual-level approaches behave about the same way: neither the design-based nor the model-based approach produces valid statistical inferences until the number of clusters is at least 30. This makes sense: both approaches rely on central limit theorems so that a Normal law can describe the distribution of the test statistic under the null hypothesis. The cluster-level approach is always valid, but sometimes produces overly large confidence intervals (when the number of clusters is small). When the number of clusters is large (say, 100), then all approaches are equivalent in terms of their error rates. Designs with few clusters should consider either the cluster-level approach using the normal approximation shown here or even direct permutation based approaches to statistical inference.


# 9 Power analysis for clustered designs

We want designs that are likely to reject hypotheses inconsistent with the data, and unlikely to reject hypotheses consistent with the data. We’ve seen that the assumptions required for the validity of common tests (typically, large numbers of observations, or large quantities of information in general) are challenged by clustered designs, and the tests which account for clustering can be invalid if the number of clusters is small (or information is low at the cluster level in general). We’ve also seen that we can produce valid statistical tests for hypotheses about the average treatment effect using either Robust Clustered Standard Errors (RCSE), multilevel models or using the cluster-level approach described by @hansen2008covariate, and that pair-matching can drastically minimize bias in designs with unequal cluster sizes.

The most important rule regarding the statistical power of clustered designs is that more small clusters are better than fewer larger ones. This can be demonstrated through simulated experiments. Generally speaking, the most flexible way to evaluate the power of a design is through simulation, as it allows for complex clustering and blocking schemes, and can incorporate covariates. In the following we use the OLS estimator with Robust Clustered Standard Errors, in order to save on computation time, but the same analysis can be achieved using any estimator and test statistic.

<style>
div.hidecode + pre {display: none}
</style>
<script>
doclick=function(e){
e.nextSibling.nextSibling.style.display="block";
}
</script>

<div class="hidecode" onclick="doclick(this);">[Click to show code]</div>
```{r,cache = T}

# A function to test the null hypothesis and the true hypothesis
test_H0_and_Htrue <- function(J = J,n = n,treatment_effect = treatment_effect,ICC = ICC) {
  # Make data:
  data <- make_clustered_data(
    J = J,
                      n = n,
                      treatment_effect = treatment_effect,
    ICC = ICC
  )
  linear_fit <- lm(Y ~ Z,data = data)

  RCSE_CI <- confint_HC(
    coefficients = coef(linear_fit),
    vcov_mat = RCSE(
      model = linear_fit,
                             cluster = data$j,
      return_cov = TRUE
    ),
    df = linear_fit$df.residual
  )["Z", ]

  # Zero should not be in this CI very often as the null of 0 is false here
  correct_reject <- !((0 >= RCSE_CI[1]) & (0 <= RCSE_CI[2]))

  # Test null of true taus (first attempt is use true, second is to make 0 true)
  # Reassign village level treatment so that Y is indep of Z --- so true effect is 0

  data$Z_new <- ifelse(data$j %in% sample(1:J, max(J)/2), 1, 0)

  linear_fit_true <-lm(Y ~ Z_new,data = data)

  RCSE_CI_true <- confint_HC(
    coefficients = coef(linear_fit_true),
    vcov_mat = RCSE(
      model = linear_fit_true,
                                             cluster = data$j,
      return_cov = TRUE
    ),
    df = linear_fit$df.residual
  )["Z_new", ]

  # Zero should be in this CI very often as the null of 0 is true here
  false_positive <-  !((0 >= RCSE_CI_true[1]) & (0 <= RCSE_CI_true[2]))

  return(c(
    correct_reject = correct_reject,
    false_positive = false_positive
  ))
}
```

We can now analyze how power is affected when the number of clusters and the size of clusters vary, holding the ICC constant at .01 and the treatment effect constant at .2.
We look both at the power --- how often we correctly reject the null of no effect when there really is one -
as well as at the error --- how often we incorrectly reject the null of no effect when there really isn't an effect. Typically we want the power to be around .8 and the error rate to
be around .5 (when using a 95% confidence level).

<style>
div.hidecode + pre {display: none}
</style>
<script>
doclick=function(e){
e.nextSibling.nextSibling.style.display="block";
}
</script>

<div class="hidecode" onclick="doclick(this);">[Click to show code]</div>
```{r,cache=T}
# The numbers of clusters we will consider
Js <- c(8,20,40,80,160,320)
# The cluster sizes we will consider
n_js <- c(8,20,40,80,160,320)

# Create an empty matrix to store the results
# The first stores the power and the second stores the error
power_J_n <- error_J_n <- matrix(
  data = NA,
  nrow = length(Js),
  ncol = length(n_js),
  dimnames = list(
    paste("J =",Js),
    paste("n_j =",n_js)
  )
)
# Set the number of simulations
sims <- 100

# Loop through the different cluster numbers
for( j in 1:length(Js)){
  # Loop through the different cluster sizes
  for(n in 1:length(n_js)){
    # Estimate the power and error rate for each combination
    test_sims <- replicate(
      n = sims,
      expr = test_H0_and_Htrue(
        J = Js[j],
                                       n = n_js[n],
                                       treatment_effect = .25,
        ICC = .01
      )
    )
    power_error <- rowMeans(test_sims)
    # Store them in the matrices
    power_J_n[j,n] <- power_error[1]
    error_J_n[j,n] <- power_error[2]
  }
}

# Plot the power under the different scenarios
matplot(power_J_n, type = c("b"),pch=1,axes = F,ylim = c(0,1.5),ylab = "power")
axis(side = 1,labels = rownames(power_J_n),at = 1:6)
axis(side = 2,at = seq(0,1,.25))
abline(h=.8)
legend("top", legend = colnames(power_J_n),col = 1:6 ,pch=1,horiz = TRUE)
```

<style>
div.hidecode + pre {display: none}
</style>
<script>
doclick=function(e){
e.nextSibling.nextSibling.style.display="block";
}
</script>

<div class="hidecode" onclick="doclick(this);">[Click to show code]</div>
```{r,cache=T}
# Plot the error rate under the different scenarios
matplot(error_J_n, type = c("b"),pch=1,axes = F,ylim = c(0,.5),ylab = "error rate")
axis(side = 1,labels = rownames(error_J_n),at = 1:6)
axis(side = 2,at = seq(0,1,.25))
abline(h=.05)
legend("top", legend = colnames(error_J_n),col = 1:6 ,pch=1,horiz = TRUE)
```

We see that power is always low when the number of clusters is low, regardless of how
large the clusters are. Even with huge clusters (with 320 units each), the statistical
power of the study is still relatively low when the number of clusters is 8.
Similarly, it takes a large number of clusters to power a study with small clusters: although it is sufficient to have many clusters in order to power an experiment, irrespective of cluster size, power increases much more rapidly when the clusters are larger.
Note also that while error rates appear systematically related to the number of clusters,
the same is not true for cluster sizes.

Next, we can evaluate how the intra-cluster correlation affects power. We will hold
the structure of the sample size constant at $J=80,n_j=80$ and $J=160,n_j=160$,
and compare across a range of ICC values, from low (.01) to high (.6).

<style>
div.hidecode + pre {display: none}
</style>
<script>
doclick=function(e){
e.nextSibling.nextSibling.style.display="block";
}
</script>

<div class="hidecode" onclick="doclick(this);">[Click to show code]</div>
```{r,cache = T}
J_njs <- c(80,160)
ICCs <- seq(0,.6,.1)+c(.01,0,0,0,0,0,0)
power_ICC <- error_ICC <- matrix(
  data = NA,
  nrow = length(ICCs),
  ncol = length(J_njs),
  dimnames = list(
    paste("ICC =",ICCs),
    paste("J =",J_njs,"n_j = ",J_njs)
  )
)
# Set the number of simulations
sims <- 100

# Loop through the different ICCs
for( i in 1:length(ICCs)){
  # Loop through the different cluster sizes
  for(j in 1:length(J_njs)){
    # Estimate the power and error rate for each combination
    test_sims <- replicate(
      n = sims,
      expr = test_H0_and_Htrue(
        J = J_njs[j],
                                       n = J_njs[j],
                                       treatment_effect = .25,
        ICC = ICCs[i]
      )
    )
    power_error <- rowMeans(test_sims)
    # Store them in the matrices
    power_ICC[i,j] <- power_error[1]
    error_ICC[i,j] <- power_error[2]
  }
}

# Plot the power under the different scenarios
matplot(power_ICC, type = c("b"),pch=1,axes = F,ylim = c(0,1.5),ylab = "power (high ICC)")
axis(side = 1,labels = rownames(power_ICC),at = 1:7)
axis(side = 2,at = seq(0,1,.25))
abline(h=.8)
legend("top", legend = colnames(power_ICC),col = 1:2 ,pch=1,horiz = TRUE)
```

<style>
div.hidecode + pre {display: none}
</style>
<script>
doclick=function(e){
e.nextSibling.nextSibling.style.display="block";
}
</script>

<div class="hidecode" onclick="doclick(this);">[Click to show code]</div>
```{r,cache=T}
# Plot the error rate under the different scenarios
matplot(error_ICC, type = c("b"),pch=1,axes = F,ylim = c(0,.5),ylab = "error rate")
axis(side = 1,labels = rownames(error_ICC),at = 1:7)
axis(side = 2,at = seq(0,1,.25))
abline(h=.05)
legend("top", legend = colnames(error_ICC),col = 1:2 ,pch=1,horiz = TRUE)
```

As this example illustrates, high ICC can severely diminish the
statistical power of the study, even with many large clusters.

# 10 How to check balance in clustered designs

Randomization checks in clustered designs follow the same form as the preceding discussion. A valid test for a treatment effect is a valid test for placebo or covariate balance. The only difference from our preceding discussion is that one uses a background covariate or baseline outcome --- some variable putatively uninfluenced by the treatment --- in place of the outcome itself. So, randomization tests with small numbers of clusters may be too quick to declare an experiment ill-randomized if the analyst is not aware of the methods of error-rate analysis that we described above.

One new problem does arise in the context of randomization tests. Often one has many covariates which could be used to detect unlucky imbalances or field problems with the randomization itself. And, if one uses hypothesis tests, then, of course, a valid test which encourages us to declare “imbalance” when $p<.05$ would do so falsely for one in every twenty variables tested. For this reason, we recommend using one-by-one testing as an exploratory tool and using omnibus tests (like the Hotelling T-test or an F-test or the @hansen2008covariate $d^2$ test), which can combine information across many dependent tests into one test statistic to make balance tests directly. However, these tests must account for the clustered nature of the design: a simple F-test without accounting for the clustered-design will likely mislead an analyst into declaring a design unbalanced and perhaps charging the field staff with a randomization failure.

Since cluster randomized experiments tend to have cluster-level covariates (say, village size, etc..) balance checks at the cluster level make sense and do not require explicit changes to account for clustered-assignment. @hansen2008covariate develop such a test and provide software to implement it. So, for example, if we had 10 covariates measured at the village level, and we had a large number of villages we could assess an omnibus balance hypothesis using this design-based but large-sample tool.

Here we show only the omnibus test results. The one-by-one assessments that make up the omnibus test are also available in the `balance_test` object. Here, the omnibus test tell us that we have little information against the null that these observations arose from a randomized study.

<style>
div.hidecode + pre {display: none}
</style>
<script>
doclick=function(e){
e.nextSibling.nextSibling.style.display="block";
}
</script>

<div class="hidecode" onclick="doclick(this);">[Click to show code]</div>
```{r,cache = T}
library(RItools)
options(digits=3)

# Make a village level dataset
villages <- aggregate(pretend_data,by = list(village = pretend_data$j), mean)

# Generate 10 fake covariates
set.seed(12345)
villages[paste("x",1:10,sep="")] <- matrix(rnorm(nrow(villages)*10), ncol=10)
balance_formula <- reformulate(paste("x",1:10,sep=""), response="Z")
# Do a design-based, large sample balance test
balance_test <-xBalance(balance_formula,
                        strata=list(noblocks=NULL),
                        data=villages,
  report = c(
    "std.diffs", "z.scores", "adj.means",
    "adj.mean.diffs", "chisquare.test", "p.values"
  )
)

# The results of the 1-by-1 balance tests
kable(round(balance_test$results,2),align = "c")
```

<style>
div.hidecode + pre {display: none}
</style>
<script>
doclick=function(e){
e.nextSibling.nextSibling.style.display="block";
}
</script>

<div class="hidecode" onclick="doclick(this);">[Click to show code]</div>
```{r,cache=T}
# The overall omnibus p-value
kable(round(balance_test$overall,2),align = "c")
```

In this case, we cannot reject the omnibus hypotheses of balance even though, as we expected, we have a few covariates with falsely low $p$-values. One way to interpret this omnibus result is to say that such imbalances on a few covariates would not appreciably change any statistical inferences we make about treatment effects as long as these covariates did not strongly predict outcomes in the control group. Alternatively, we could say that any large experiment can tolerant chance imbalance on a few covariates (no more than roughly 5% if we are using $\alpha=.05$ as our threshold to reject hypotheses).


# References
